Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-05-15 Manuscript/manuscript.Rmd
edits to results section; added full results including tables and figures; some place holders for figures

To investigate whether we could alleviate resolution limitations of conventional TAD and chromatin  loop calling tools, we developed *preciseTAD*, an algorithm that leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base pair (bp) resolution. To demonstrate, we opted to implement an optimally built whole genome RF model. That is, a random forest model built on the optimal combination of data resolution (binning), predictor type (feature space), re-sampling technique (class balancing correction), and annotations (most informative genomic elements) [??? Write how your vector Y and annotation matrices change;!!!added ???You describe whole-genome model. It should have been described early on. Here, describe chromosome-specific bp-resolution model]. Here, $Y$ was of length $N$ where $N=n_1+n_2+\cdots+n_{22}$, the total number of bins across each chromosome. The data matrix $\textbf{X}$ had dimensions $N \times p_s$, where $p_s \subset P$, a subset of the most predictive genomic elements. The whole genome RF was tuned sequentially. First, *mtry*=$\sqrt{p}$ and *nodesize*=1, then we allowed *ntree* to range from 10 to 1000 by values of 10, and determined where the balanced accuracies converged, say at $t$. Next, we set *mtry*=$\sqrt{p}$ and *ntree*=$t$ and allowed *nodesize* to range from 1 to 21 by values of 2, again evaluating where performances converged. <!--Given that each chromosome is composed of tens, in some cases hundreds, of millions of base pairs,--> We then used the same whole genome RF model to predict boundaries from chromosome-specific bp resolution data.

@@

Additionally, we compared the $log_2$ distance between TAD boundaries and the same top predictive transcription factor binding sites using Wilcoxon Rank-Sum test. The alpha level of significance was set to 0.05.

@@

where A and B represent genomic regions created by flanking called and predicted boundaries. That is, the set of shared boundaries divided by the total number of unique boundaries.

@@

Here, we opted to implement preciseTAD using an optimally built whole genome random forest model. That is, a random forest model built on 5 kb defined genomic bins, with random under-sampling, and distance-type predictors for the top informative TFBS (CTCF, RAD21, SMC3, and ZNF143). The reason for this was because of the improved performance that was exhibited for a whole genome random forest compared to aggregated chromosome-specific models (Supplementary Figure 7). When tuning the full genome RF we found that performances converged when *mtry*=2, *ntree*=100, and *nodesize*=???.

???stopped here; waiting to predicted tad boundaries results

<!--
We first used ROC analysis to obtain an optimal probability threshold for defining potential predicted TAD boundaries. We reported an average threshold of ??? (SD=???) among the 21 autosomal chromosomes that were analyzed (mean=???, SD=??? for K562). We referred to collections of chromosome-specific base pairs that exceeded this threshold when used to validate our *TADRF* model as *predicted TAD boundary areas* (PTBA), given that collections of adjacent base pairs tended to exhibit similar prediction profiles. We found that there were a total of ??? PTBA, containing an average of ??? base pairs each (??? PTBA for K562; mean base pairs=???, SD=???). *preciseTAD* then implements complete linkage hierarchical clustering on the dissimilarity matrix constructed from the pairwise distances (in base pairs) among the midpoints of each chromsome-specific PTBA. Using the silhouette score (see methods), we reported a total of k=??? chromosome-specific clusters (k=??? for K562). We referred to these *k* clusters as *predicted TAD boundary regions* (PTBR). That is, each PTBR contained the midpoints of multiple PTBA. Lastly, given each PTBR, *preciseTAD* uses CLARA partitioning to obtain the most representative midpoint. We referred to this as the *predicted TAD boundary point* (PTBP). These PTBP were identified as the most representative base pair among each PTBR. *preciseTAD* was able to identify a total of ??? PTBP across each chromosome. We reported less predicted TAD boundaries than ARROWHEAD at 10 kb (Table 1). This is attributed to ARROWHEAD's inherent inflation of called TAD boundaries, especially near fuzzy areas along the diagonal of Hi-C contact matrices (Figure 5A). Likewise, under the framework of *preciseTAD*, any two TADs are inherently demarcated by only one predicted TAD boundary, leading to comparatively less TAD boundaries. We also find that predicted TAD boundaries identified by *preciseTAD* were much more likely to demarcate high concentrations of inter-chromosomal contacts among Hi-C contact matrices (Figure 5A). [???include a result that demonstrates this; discuss in meeting]
-->

<!--We were able to identify regions where the model predicted the base pair to be a TAD boundary with probability equal to 1 (Figure 5A). [??? Where is this figure? It seemed important; !!!added] By clustering the regions (see Methods), we were able to predict a comparable number of TAD boundary regions as called by the ARROWHEAD TAD caller. However, only 24% and 29% of predicted and called (and 10 kb flanked) TAD boundaries overlapped for the GM12878 and K562 cell line, respectively, suggesting differing locations in TAD boundaries throughout the linear genome (Supplementary Figure 5). [??? Need further drill down, as we discussed; !!!added distance to annotations figure in next paragraph]

For each predicted TAD boundary region, we were able to identify the summit locations of TAD boundary location by performing partitioning around medoids (see Methods). To compare the summit locations of predicted and called TAD boundaries, we evaluated the signal strength of ChIP-seq peak regions around a 10 kb flanking region. We found that the signal strength for the CTCF, RAD21, SMC3, and ZNF143 [??? Make all in this order!; !!!fixed] TFBS peaked at our predicted TAD boudaries, while was poorly defined around the called TAD boundaries (Figure 5B; Supplementary Figure 6).  Thus, our model is able to alleviate resolution limitations and precisely locate TAD boundaries that correspond to the highest signal strength of known TAD boundary-forming TFBSs. Furthermore, when compared to called TAD boundaries, the boundaries that we predicted with our model were found to be statistically significantly closer to the top ranked TFBS, suggesting stronger enrichment profiles near predicted TAD boundaries (Supplementary Figure 7).

![**Predicted TAD boundaries are enriched in known TAD boundary-forming TFBSs.** (A) Visualizing locations of predicted and called TAD boundaries (chr 22: 17389000-18011000; GM12878). The upper triangular contact map shows the normalized number of contacts within the specified window using the JUICER toolbox.The blue vertical lines represent called boundaries identified by ARROWHEAD, with the corresponding called TADs outlined in blue on the upper diagonal contact matrix above. The black line represents the predicted probability of each base pair being a TAD boundary when validated by *TADRF*. Collections of red vertical lines represent base pairs whose predictive probability exceed the ROC threshold, labeled as *predicted TAD boundary areas* (PTBA). The midpoints of spatially similar PTBA were clustered together into *predicted TAD boundary regions* (PTBR), denoted by the yellow regions along the x-axis. The green vertical lines represent the *predicted TAD boundary points* (PTBA) identified by *preciseTAD*, with the corresponding predicted TADs outlined above in green.](Figures/visualizing_ptbp_and_ctbp.png) 
-->

@@

**Supplementary Figure 7. Whole genome random forest model outperform chromosome-specific random forest models across resolutions.** Random forest models built on ARROWHEAD boundaries using 5 kb defined genomic bins, with random under-sampling, and distance-type predictors for the top informative TFBS (CTCF, RAD21, SMC3, and ZNF143) applied to the entire genome (A) exhibit higher balanced accuracies than their chromosome specific counterparts (B) across each resolution respectively.

>>>>

To investigate whether we could alleviate resolution limitations of conventional TAD and chromatin  loop calling tools, we developed *preciseTAD*, an algorithm that leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base pair (bp) resolution. To demonstrate, we opted to implement an optimally built whole genome RF model. That is, a random forest model built on the optimal combination of data resolution (binning), predictor type (feature space), re-sampling technique (class balancing correction), and annotations (most informative genomic elements) [??? Write how your vector Y and annotation matrices change;!!!added ???You describe whole-genome model. It should have been described early on. Here, describe chromosome-specific bp-resolution model]. Here, $Y$ was of length $N$ where $N=n_1+n_2+\cdots+n_{22}$, the total number of bins across each chromosome. The data matrix $\textbf{X}$ had dimensions $N \times p_s$, where $p_s \subset P$, a subset of the most predictive genomic elements. The *nodesize* and *ntree* parameters were tuned sequentially using a grid search across values from ???-??? by ??? and from ???-??? by ??? for each, respectively. We then used the same optimized whole genome RF model to predict boundaries from chromosome-specific bp resolution data.

@@

Additionally, we compared the median $log_2$ distances between TAD boundaries and the same top predictive transcription factor binding sites using Wilcoxon Rank-Sum test.

@@

where A and B represent genomic regions created by flanking called and predicted boundaries. That is, the set of shared boundaries divided by the total number of unique boundaries. Wilcoxon rank sum tests were used to compare chromosome-specific jaccard indices across cell lines, between *preciseTAD* boundaries and both ARROWHEAD and PEAKACHU boundaries respectively. The alpha level of significance was set to 0.05 for all statistical tests.

@@

At this point, we opted to implement preciseTAD using an optimally built whole genome random forest model. That is, a random forest model built on 5 kb defined genomic bins, with random under-sampling, and distance-type predictors for the top informative TFBS (CTCF, RAD21, SMC3, and ZNF143). The reason for this was because of the improved performance that was exhibited for a whole genome random forest compared to aggregated chromosome-specific models (Supplementary Figure 5). When tuning the full genome RF we found that performances converged when *mtry*=2, *ntree*=250, and *nodesize*=??? (Supplementary Figures 6-7).

When built using ARROWHEAD, at 5 kb, *preciseTAD* was able to predict a total of ??? and ??? boundaries for GM12878 and K562 cell lines respectively. We reported less predicted TAD boundaries than ARROWHEAD (Table 1, Supplementary Table 3). This is attributed to ARROWHEAD's inherent inflation of called TAD boundaries, especially near fuzzy areas along the diagonal of Hi-C contact matrices (Figure 5A). Likewise, under the framework of *preciseTAD*, any two TADs are inherently demarcated by only one predicted TAD boundary, leading to comparatively less TAD boundaries. Visually speaking, we also found that predicted TAD boundaries identified by *preciseTAD* were much more likely to demarcate high concentrations of inter-chromosomal contacts among Hi-C contact matrices (Figure 5A). When built using PEAKACHU loop boundaries, *preciseTAD* was able to predict a total ??? and ??? boundaries for GM12878 and K562 cell lines respectively. This can be attributed to the fact that PEAKACHU's is built using soley CTCF sites, while *preciseTAD* also incorporates information from other architectural proteins including RAD21, SMC3, and ZNF143.

Next, we evaluated the biological significance of *preciseTAD* boundaries. Enriched heatmaps indicated that our predicted boundaries were more enriched for CTCF sites than either ARROWHEAD or PEAKACHU boundaries alone (Figure 5 B-C. Likewise, *preciseTAD* boundaries were also more enriched for other molecular drivers of 3D chromatin including RAD21, SMC3, and ZNF143 for both GM12878 and K562 cell lines (Supplementary Figures 8-9). Furthermore, when compared to called TAD or loop boundaries, *preciseTAD* were found to be statistically significantly closer to the same cell line-specific top ranked TFBS (Figure 5D, Supplementary Figure 10-11).

![**preciseTAD boundaries are enriched in CTCF sites.** (A) Visualizing locations of predicted vs called ARROWHEAD TAD boundaries on GM12878 (chr 21: 15150000-17270000). The upper triangular contact map shows the normalized number of contacts within the specified window using the JUICER toolbox. The blue vertical lines represent boundaries identified by ARROWHEAD, with the corresponding called TADs outlined in blue on the upper diagonal contact matrix above. The black line represents the predicted probability of each base pair being a TAD boundary when validated by our random forest. The green vertical lines represent the *preciseTAD boundary points* (PTBP) predicted by *preciseTAD*, with the corresponding demarcated TADs outlined above in green. Below is a zoomed in portion of the genome showing a cluster of base pairs with high probability representing a single *preciseTAD boundary region (PTBR) colored in yellow. (B) Enriched heatmaps comparing the enrichment of CTCF sites around 5 kb flanked preciseTAD and ARROWHEAD TAD boundaries respectively. (C) Enriched heatmaps comparing the enrichment of CTCF sites around 10 kb flanked preciseTAD and PEAKACHU loop boundaries respectively. (D) $Log_{2}$ genomic distances from boundaries to nearest CTCF site between preciseTAD boundaries and both ARROWHEAD and PEAKACHU boundaries, respectively.](Figures/figure5_enrichment_and_log2dist_ctcf.png)

## *preciseTAD* boundaries are conserved across cell lines

Subsequent comparisons of *preciseTAD* with both ARROWHEAD and PEAKACHU revealed that our predicted boundaries were more conserved between the GM12878 and K562 cell lines. To assess conservation, we evaluated the overlap between cell line specific boundaries after flanking by the resolution use to make predictions. Boundaries involving ARROWHEAD were flanked by 5 kb, while boundaries involving PEAKACHU were flanked by 10 kb. We found that a significant proportion of flanked boundaries predicted by *preciseTAD* overlapped between GM12878 and K562. Specifically, 45%??? and 61%??? of *preciseTAD* boundaries were conserved across cell lines when built on ARROWHEAD and PEAKACHU boundaries respectively (Figure 6A). Additionally, median jaccard index (MJI) was found to be 0.472 (IQR=0.441-0.503; p=???) and 0.647 (IQR=0.646-0.647; p=???) for *preciseTAD* predicted TAD and loop boundaries (Supplementary Table 5). Comparatively, only 25% (MJI=0.284; IQR=0.268-0.301) and 49% (MJI=0.509; IQR=0.503-0.515) of called/predicted ARROWHEAD and PEAKACHU boundaries were conserved (Figure 6A, Supplementary Table 5).

Expectedly, the majority of flanked TAD boundaries predicted by *preciseTAD* represented a subset of the larger group of predicted loop boundaries. This makes sense as the boundaries that demarcate chromatin loops make up a broader class of boundaries, with TAD boundaries representing the next level in 3D chromatin hierarchy, followed by A/B compartments. For GM12878, 83% of predicted TAD and loop boundaries overlapped with eachother, while only 3% of predicted TAD boundaries were not contained in the set of flank-predicted loop boundaries (Figure 6B). Similarly, for K562, 75% overlap occured between predicted TAD and loop boundaries, with only 4% of predicted TAD boundaries remaining outside of the set of flank-predicted loop boundaries (Figure 6B).

![**preciseTAD boundaries are conserved between the GM12878 and K562 cell lines.** (A) Venn diagrams illustrating the percentage of overlap between GM12878- and K562-specific predicted and called boundaries. (B) Venn diagrams illustrating the nested structure of predicted TAD and loop boundaries identified by *preciseTAD* for the GM12878 and K562 cell lines. Note, predicted and called TAD boundaries were flanked by 5 kb, while predicted and called loop boundaries were flanked by 10 kb prior to calculating overlaps to reflect the resolution that was used to predict each respectively.](Figures/figure6_venn_diagrams.png)

@@

**Supplementary Figure 5. Whole genome random forest model outperform chromosome-specific random forest models across resolutions.** Random forest models built on ARROWHEAD boundaries using 5 kb defined genomic bins, with random under-sampling, and distance-type predictors for the top informative TFBS (CTCF, RAD21, SMC3, and ZNF143) applied to the entire genome (A) exhibit higher balanced accuracies than their chromosome specific counterparts (B) across each resolution respectively.

**Supplementary Figure 6. Whole genome RF performances converge when ntree=250.** Scatter plot of cross-validated balanced accuracies (BA) for random forests built using both ARROWHEAD (circles) and PEAKACHU (triangles) boundaries, with random under-sampling, distance-type predictors (CTCF, RAD21, SMC3, and ZNF143), over the entire binned genome, for GM12878 (red) and K562 (blue) respectively. For the ARROWHEAD model, 5kb bins were used, while 10kb bins were used for the PEAKACHU model. Performances converged after trees of size 250, marked by the dashed line.

**Supplementary Figure 7. Whole genome RF performances converge when nodesize=???.** Scatter plot of cross-validated balanced accuracies (BA) for random forests built using both ARROWHEAD (circles) and PEAKACHU (triangles) boundaries, with random under-sampling, distance-type predictors (CTCF, RAD21, SMC3, and ZNF143), over the entire binned genome, for GM12878 (red) and K562 (blue) respectively. For the ARROWHEAD model, 5kb bins were used, while 10kb bins were used for the PEAKACHU model. Performances converged after nodesize values of ???, marked by the dashed line.

**Supplementary Figure 8. preciseTAD boundaries are more enriched for GM12878-specific SMC3, RAD21, and ZNF143 sites.** Enriched heatmaps comparing preciseTAD boundaries with both (A) ARROWHEAD and (B) PEAKACHU boundaries for the GM12878 cell line.

**Supplementary Figure 9.preciseTAD boundaries are more enriched for K562-specific CTCF, SMC3, RAD21, and ZNF143 sites.** Enriched heatmaps comparing preciseTAD boundaries with both (A) ARROWHEAD and (B) PEAKACHU boundaries for the K562 cell line.

**Supplementary Figure 10. preciseTAD boundaries are spatially closer to GM12878-specific SMC3, RAD21, and ZNF143 sites.** Boxplots comparing the $log_{2}$ genomic distances from each preciseTAD boundary (green) and called boundary (either ARROWHEAD or PEAKACHU; blue) to the nearest SMC3, RAD21, and ZNF143 sites on the linear genome for GM12878.

**Supplementary Figure 11. preciseTAD boundaries are spatially closer to K572-specific CTCF, SMC3, RAD21, and ZNF143 sites.** Boxplots comparing the $log_{2}$ genomic distances from each preciseTAD boundary (green) and called boundary (either ARROWHEAD or PEAKACHU; blue) to the nearest CTCF, SMC3, RAD21, and ZNF143 sites on the linear genome for K562.

**Supplementary Table 5.** Median chromosome-specific Jaccard indices (IQR) for called and *preciseTAD* predicted TAD (ARROWHEAD) and loop (PEAKACHU) boundaries between the GM12878 and K562 cell lines. The p-value is from a nonparametric Wilcoxon-rank sum test. Note, predicted and called *TAD* boundaries were flanked by 5 kb, while predicted and called *loop* boundaries were flanked by 10 kb prior to calculating Jaccard indices to reflect the resolution that was used to predict each respectively.
